Search for Spatially Extended Fermi-LAT Sources Using Two Years of Flight Data

### 2011-05-21 extended_catalog/paper/extended_catalog.tex
better discussion

\documentclass{emulateapj}

\usepackage{fullpage}
\usepackage{graphicx}

@@

\usepackage{amsmath}

@@

\begin{abstract}
We present a new method for analyzing spatially extended sources with
the Fermi Large Area Telescope (LAT), the primary science instrument
on the {\em Fermi Gamma-ray Space Telescope (Fermi)}.  We provide a
series of Monte Carlo studies that were done to validate this tool.
We then apply this tool to test all sources in the 2FGL Two Year Catalog
for extension.\cite{2FGL} We then report on the discovery of XXX new
extended sources.
\end{abstract}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the
Fermi Gamma-Ray Space Telescope that has been surveying the gamma ray
sky since June 2008.  Fermi is particuarly well suited for surveying
the entire Gamma-ray sky since it has a broad energy coverage (20MeV
and 300GeV), a wide field of view (~2.4 sr) and large effective area
($\approx 8000 \text{cm}^2$ at > 1GeV).

@@

Resolving spatially extended GeV sources is important for studying a
variety of science topics. The most important use of spatial morphology
is in source identification. Many of the sources found in GeV sources

Nevertheless, morphological studies of sources in the GeV energy range using
the Fermi LAT is challenging primarily because of the significantly energy
dependent PSF and because of systematic errors in the
galactic diffuse emission. The Fermi PSF varies from over $XXXXX^\circ$
at 100MeV to $XXXXX^\circ$ at 100GeV and so the higher energy photons are

studying supernova remnants, 
pulsar wind nebulae, 
 * galaxies,  
 * potentially dark matter or new physics.

@@

Using 11 months of data, the LAT produced the best-resolved survey of the sky
in the 100 MeV to 100 GeV energy range\cite{2FGL Paper}. The

@@

Performing morphological studies of sources in the GeV energy range using
the Fermi LAT is challenging primarily because of the significantly energy
dependent PSF and because of systematic errors in the
galactic diffuse emission. The Fermi PSF varies from over $XXXXX^\circ$
at 100MeV to $XXXXX^\circ$ at 100GeV and so the higher energy photons are
significantly more important for studying the shape of sources. One could
ideally look just at the highest energy photons, but the trade off is that
for typical sources many more source events are expected at low energy.
For this reason, it is especially important to use as much of the data
as possible.

@@

In principle, the entire task can be done using the standard science
tools package {\em gtlike}\cite{Science-Tools-gtlike}. But {\em gtlike} is
only able to fit the spectral parameters of the source, so the extension
fitting would have to be done as an external loop. Typically, the run time
for a binned source analysis using {\em gtlike} is several hours so this loop
would be especially time consuming. In principle, parallel computing could
ease this burdon, but parallel fitting algorithms are rather untested
and complex. Furthermore, the machinery required to do something like
this would necessarily be rather complex. What has typically been done by
the Fermi team to study extended sources is to fix the assumed position
of the extended source, either by our knowledge of a source from other
wavelenghts, or by fitting it as a point source. Then, a profile of the
likelihood as a function of extension is developed using {\em gtlike}
and the maximum of this profile. This approach is not optimal because it
needs not be the case that the best fit position of a point source would
be the same as the best fit position of an extended source. Furthermore,
by not fully maximizing the likelihood, this method is not as significant to
extension, especially of faint sources. And finally, since this approach
is so computationally intensive, no large scale Monte Carlo effort has been
launched to validate the method.

Instead, an alternate approach was developed for performing
a morphological analysis. An alternate maximum likelihood fitting
package called {\em pointlike} has been developed by the Fermi
collaboration and we have expanded its functionality to fully support
analyzing spatially extended sources. {\em pointlike} is most thoroughly
described in Matthew Kerr's Ph.D Thesis\cite{Matthew_kerr_phd_thesis}
but the important aspects will be summarized below. What differentiates
{\em pointlike} most significantly from {\em gtlike} is its emphasis
on speed by introducing approximations into the calculation of the
likelihood function while at the same time trying to reduce any
numerical errors that may be introduced.  Like binned {\em gtlike},
{\em pointlike} bins the sky both in position and energy.  Unlike {\em
gtlike}, pointlike relies on the {\em healpix} representation of spatial
bins\cite{healpix_paper} and scales the size of each spatial bin to be
smaller than the PSF. This is convenient because it no
significant information is lost due to binning at all energies and it is
always a good approximation that the model predicted counts are uniform
across a spatial bin. Furthermore, {\em pointlike} uses a sparse matrix
representation of the spatial bins where only the bins with counts in them
are looped over. The only downside to this is that the integral of the
model predicted counts must be independently calculated.  Effectively,
{\em pointlike} efficiently interpolates from a totally binned analysis
at low energy (where the resolution is poor and there are many counts)
to an unbinned analysis at high energy (where the resolution is very
good but there are few counts). This approach provides a significant
savings in time while introducing only small numerical error.

We have modified {\em pointlike} to support fitting spatially extended
sources.  To do this, we introduced a new class of diffuse sources where
the spatial and spectral information completely decouple and where
the spatial shape is parameters by simple objects (Disks, Gaussian).
The code consistently convolves the extended source shape with the point
spread function (as a function of energy) and integrates the spectrum over energy
to computed model predicted counts.

Finally, {\em pointlike} allows the spatial parameters of an extended
source to be simulataneoulsy fit. {\em pointlike} uses the Minuit fitting
library to perform the maximum likelihood maximization of the position
and extension of an extended source\cite{Minuit reference}.  At each
step in the fit of the shape, the spectral parameters of the entire sky
model are reoptimized. To avoid projection effects, what is directly
fit is not the longitude and latitude of the source but instead
the displacement in a rotated reference frame. The covariance matrix is
then computed and inverted to estimate the statistical errors on the
position and extension.

{\em pointlike} optimizes the analysis of extended sources in two ways.
First, for extended sources which are radially symmetric, the convolution
of the source shape with our parameterization of the PSF can be performed
semi-analytically. Generally, at a given energy the observed 
photon distribution is
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') d A' 
\end{equation}
where $I_\text{src}$ is the extended soruce shape and the PSF is parameterized by
\begin{equation}
  \text{PSF}(\vec r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma}
\end{equation}
Here, $\sigma$ and $\gamma$ are parameterize the $PSF$.
If the extended source is radially symmetric, then
$I_\text{src} (\vec r') = I_\text{src} (r')$ and the angular part of the
integral can be done analytically, leaving a simpler one dimensional
integral:
\begin{multline}
  \text{PDF}(u)= \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma \\
  \times ~_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right)
\end{multline}
Not only is this convolution formula faster, but it has the advantage
that it more accurately reduces to the actual PSF shape as the source's
size decreases.  Decreasing the numerical difference between a small extended
source and the PSF is important when searching for smaller extended sources
and when determining the statistical significance of a detection.

@@

To calculate the statistical significance of the extension of a source,
one can define a parameter $\text{TS}_\text{ext}$ as twice the increase
in log likelihood between fitting a source with a
spatially extended hypothesis and a point hypothesis. 
So $\text{TS}_\text{ext}=2\times(LL_\text{ext}-LL_\text{ps})$.

In going from fitting a point source to fitting a simple extended source such as
a disk, the source's width is the only degree of freedom added to the model. According to
Wilk's Theorem, the distribution of $\text{TS}_\text{ext}$ for no signal (when there is a true
point source) should follow a $\chi^2_1$ distribution with one degree of freedom\cite{Wilks_Theorem}.
Wilk's Theorem only holds when the likelihood is maximized simultaneoulsy
fitting all of the parameters of model. It is for this reason why {\em pointlike}
goes through such great effort to simultaneously fit the position, extension,
and spectral paramters of a source.

On the other hand, since the null hypothesis rests on the edge of
parameter space, when the extension is fixed to 0, Wilk's Theorem does
not formally hold\cite{Warnings about Wilk's Theorem Paper}.  One might
expect the actual distribution to be a half $\chi^2_1$ distribution with
one degree of freedom plus half of a delta function at 0:
\begin{equation}
  P(\text{TS})=\tfrac{1}{2}(\chi^2_1(\text{TS})+\delta(\text{TS})).
\end{equation}
The false detection probability can be estimated by integrating this function 
from an observed test statistic value to infinity. It is for this
particular distribution that the often quoted result holds that
$\sqrt{TS}$ is a measure of the number of $\sigma$ of the detection.

The half $\chi^2_1$ distribution was found by a monte carlo study
to hold for the special case of source finding of EGRET sources using the
maximum likelihood technique.  They were comparing the hypothesis of a
source to the hypothesis of just a background\cite{Mattox_et_All_Paper}.
Their argument for why this null hypothesis distribution was reasonable
was that half of the time, statistical underfluctuations would cause
there to be fewer fewer photons then expected where the sources is being
tested. Since the maximim likelihood machinery cannot fit negative
sources (which would lead to negative model predicted for which the
poisson likelihood is undefined), half of the tests would not have the
likelihood improved by adding a source to the model and $\text{TS}=0$.
It is for the case that the null distribution has

@@

One issue that we found when testing this analysis method is that
despite our best efforts, there was always a small discrepancy between
the likelihood value when fitting the source as a point source and when
fitting it as an extended source with a small extension.  This is to be
expected due to numerical innacuracies introduced by the convolution.
For most cases, this error is insignificant. It would not significanlty
bias any of the fit values or errors. But it would bias the distribution
of $\text{TS}_\text{ext}$ in the null hypothesis. To correct for this,
instead of computing the likelihood in the null hypothesis by fitting
a point source, a new source model was introduced which is an extended
source with its extension fixed to a very small value. $LL_\text{ps}$
is then calculated by refitting this source using the same extended
source fitting code. It was only after introducing this change that we
were able to get broad agreement between simulations and Wilk's theorem.

@@

We performed a dedicated monte carlo study to validate pointlike for
studying spatially extended sources and also to demonstrate that we can
use $\text{TS}_\text{ext}$ as a test of the statistical significance
of a source's extension.  The monte carlo study involved simulating point
sources of varying spectral models on top of a simulated background. The
simulated point source was then fit using pointlike as both an extended
source and a point source and $\text{TS}_\text{ext}$ was calculated.

Fermi LAT data was simulated using the publically avaliabable Science
Tools package {\em gtobssim}\cite{GTOBSSIM_CITATION}. The sources were
simulated with a powerlaw spectral model with 6 Were given 6 $>100\text{MeV}$
fluxes ranging from $3\times 10^{-9} ph/cm^2/s$ to $10^{-6} ph/cm^2/s$.
The sources were then simulated with spectral index ($\gamma$) values of
1.5, 2, 2.5, and 3.  These values were picked to probe a representative
range measurable values.  For particular spectral indices, only the
fluxes which would create a statistically significant source detection
($\text{TS}>25$) where selected, since we are only interested in
testing for extension sources with a significant extent. The sources were
simulated on top of an isotropic powerlaw background with a Sreekumar-like
spectrum ($>\text{100MeV}$ flux of 1.5e-5 and spectral index 2.1).\cite{Sreekumar
et al. ApJ 494 pag 523 1998}

The monte carlo simulation was perforormed over a one year time interval using
a default rocking profile and an assumed livetime fraction of 0.8.
The reconstruction was perfromed with events from an energy range of
1GeV to 100GeV using 4 energy bins per decade and using the Pass 7 Version 6
source IRFs to be consistent with the following analysis.

@@

\begin{table}
  \begin{centering}
\begin{tabular}{ | r | r | r | r | }
\hline
$\gamma$ & flux ($ph/cm^2/s$) & $N_\text{sims}$ & $<\text{TS}>$ \\
\hline
  1.5 &          $10^{-6}$ &           11405 &  90861 \\
      &  $3\times 10^{-7}$ &           11490 &  21922 \\
      &          $10^{-7}$ &           11101 &   5765 \\
      &  $3\times 10^{-8}$ &           10269 &   1262 \\
      &          $10^{-8}$ &            9941 &    306 \\
      &  $3\times 10^{-9}$ &           10427 &     61 \\
\hline
    2 &          $10^{-6}$ &           11632 &  22025 \\
      &  $3\times 10^{-7}$ &           11622 &   4886 \\
      &           $1^{-7}$ &           11115 &   1159 \\
      &  $3\times 10^{-8}$ &           10503 &    222 \\
      &           $1^{-8}$ &            9893 &     50 \\
\hline
  2.5 &          $10^{-6}$ &           11522 &   4722 \\
      &  $3\times 10^{-7}$ &           11445 &    948 \\
      &           $1^{-7}$ &           10966 &    202 \\
      &  $3\times 10^{-8}$ &            8227 &     41 \\
\hline                                                
    3 &          $10^{-6}$ &           11264 &    949 \\
      &  $3\times 10^{-7}$ &           10986 &    172 \\
      &           $1^{-7}$ &            7855 &     40 \\
\hline
\end{tabular}
\caption{List of the spectral models used to the monte carlo simulation of
{\em pointlike}.$N_\text{sims}$ is the number of monte carlo simluations
which produce a significant source detection.The final column is the
average test statistic that was found during the monte carlo fit of
the source. Only simluations which were deemed significant ($\text{TS}>25$)
were kept.}
\label{ts_ext_num_sims}
  \end{centering}
\end{table}

The results of the monte carlo simulation are shown in
figure~\ref{ts_ext}.  The cumulative density function (CDF) is plotted
against $\text{TS}_\text{ext}$ For reference, the CDF for $\chi^2_1/2$,
suggested by Wilk's Theorem, is also plotted for comparison. Overall,
there is good agreement between Wilk's theorem, although the agreement
is not perfect.  Possible reasons for departure from Wilk's theorem may
include pointlike ignoring energy dispersion, which would generally change
the shape as a function of energy, numerical innacuracies in evaluating
the convolution algorithm, or possibly problems with the fitting algorithm
getting stuck in false minimum. It is important to emphasize that if
our emperical curve lies to the left of the curve we assume to estimate
significance, we would underestimate the statistical significance of
our result. Since this is the case for most of our simluations, we are
reasonable confident that $\sqrt{\text{TS}_\text{ext}}$ can be safely
used as a measure of the statistical significance of a source's extension.

@@

We use another monte carlo study to determine the Fermi LAT's sensitivity to
spatially extended sources.

@@

% * Plots we make ...

>>>>

\documentclass[preprint]{aastex}

\usepackage{rotating}
\usepackage{amsmath}
\usepackage{graphicx}

\newcommand{\mev}{\text{MeV}}
\newcommand{\gev}{\text{GeV}}
\newcommand{\sr}{\text{sr}}
\newcommand{\cm}{\text{cm}}
\newcommand{\tsext}{\ensuremath{\text{TS}_\text{ext}}}
\newcommand{\ts}{\text{TS}}
\newcommand{\pointlike}{{\em pointlike}}
\newcommand{\gtlike}{{\em gtlike}}

@@

@@

\begin{abstract}
We present a new method for analyzing spatially extended sources with
the Fermi Large Area Telescope (LAT), the primary science instrument
on the {\em Fermi Gamma-ray Space Telescope (Fermi)}.  We provide a
series of Monte Carlo studies that were done to validate this tool.
We then calculate the Fermi LAT's sensitivity to spatially exnteded
sources.
We then apply this tool to test all sources in the 2FGL Two Year Catalog
for extension.\cite{2FGL} We then report on the discovery of XXX new
extended sources.
\end{abstract}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the
Fermi Gamma-Ray Space Telescope that has been surveying the gamma ray
sky since June 2008.  Fermi is particuarly well suited for surveying
the entire Gamma-ray sky since it has a broad energy coverage ($20\mev$
and $300\gev$), a wide field of view ($\sim 2.4 \sr$) and large effective area
($\sim 8000 \cm^2$ at $>1\gev$).

@@

studying supernova remnants, 
pulsar wind nebulae, 
galaxies,  galaxy clusters, and potentially dark matter and new physics.

Resolving spatially extended $\gev$ sources is important for studying a
variety of science topics. The most important use of spatial morphology
is in source identification. Many of the sources found in $\gev$ sources

Nevertheless, morphological studies of sources in the $\gev$ energy range using
the Fermi LAT is challenging primarily because of the significantly energy
dependent PSF and because of systematic errors in the
galactic diffuse emission. The Fermi PSF varies from over $XXXXX^\circ$
at $100\mev$ to $XXXXX^\circ$ at $100\gev$ and so the higher energy photons are

@@

Using 11 months of data, the LAT produced the best-resolved survey of the sky
in the $100\mev$ to $100\gev$ energy range\cite{2FGL Paper}. The

@@

Performing morphological studies of sources in the $\gev$ energy range using
the Fermi LAT is challenging primarily because of the significantly energy
dependent PSF and because of systematic errors in the
galactic diffuse emission. The Fermi PSF varies from over $XXXXX^\circ$
at $100\mev$ to $XXXXX^\circ$ at $100\gev$ and so the higher energy photons are
significantly more important for studying the shape of sources. One could
ideally look just at the highest energy photons, but the trade off is that
for typical sources many more source events are expected at low energy.
For this reason, it is especially important to use as much of the data
as possible.

@@

In principle, the entire task can be done using the standard science
tools package \gtlike\cite{Science-Tools-gtlike}. But \gtlike is
only able to fit the spectral parameters of the source, so the extension
fitting would have to be done as an external loop. Typically, the run time
for a binned source analysis using \gtlike is several hours so this loop
would be especially time consuming. In principle, parallel computing could
ease this burdon, but parallel fitting algorithms are rather untested
and complex. Furthermore, the machinery required to do something like
this would necessarily be rather complex. What has typically been done by
the Fermi team to study extended sources is to fix the assumed position
of the extended source, either by our knowledge of a source from other
wavelenghts, or by fitting it as a point source. Then, a profile of the
likelihood as a function of extension is developed using \gtlike
and the maximum of this profile. This approach is not optimal because it
needs not be the case that the best fit position of a point source would
be the same as the best fit position of an extended source. Furthermore,
by not fully maximizing the likelihood, this method is not as significant to
extension, especially of faint sources. And finally, since this approach
is so computationally intensive, no large scale Monte Carlo effort has been
launched to validate the method.

Instead, an alternate approach was developed for performing
a morphological analysis. An alternate maximum likelihood fitting
package called \pointlike has been developed by the Fermi
collaboration and we have expanded its functionality to fully support
analyzing spatially extended sources. \pointlike is most thoroughly
described in Matthew Kerr's Ph.D Thesis\cite{Matthew_kerr_phd_thesis}
but the important aspects will be summarized below. What differentiates
\pointlike most significantly from \gtlike is its emphasis
on speed by introducing approximations into the calculation of the
likelihood function while at the same time trying to reduce any
numerical errors that may be introduced.  Like binned \gtlike,
\pointlike bins the sky both in position and energy.  Unlike 
\gtlike, \pointlike relies on the {\em healpix} representation of spatial
bins\cite{healpix_paper} and scales the size of each spatial bin to be
smaller than the PSF. This is convenient because it no
significant information is lost due to binning at all energies and it is
always a good approximation that the model predicted counts are uniform
across a spatial bin. 
\pointlike additionaly bins the Fermi Gamma-rays by their conversion type,
which corresponds to whether the photon converted in the front or back
part of the detector. Since the PSF is smaller for front entering events,
one would expect binning each conversion type independently to improve
our sensitivity.
Furthermore, \pointlike uses a sparse matrix
representation of the spatial bins where only the bins with counts in them
are looped over. The only downside to this is that the integral of the
model predicted counts must be independently calculated.  Effectively,
\pointlike efficiently interpolates from a totally binned analysis
at low energy (where the resolution is poor and there are many counts)
to an unbinned analysis at high energy (where the resolution is very
good but there are few counts). This approach provides a significant
savings in time while introducing only small numerical error.

We have modified \pointlike to support fitting spatially extended
sources.  To do this, we introduced a new class of diffuse sources where
the spatial and spectral information completely decouple and where
the spatial shape is parameters by simple objects (Disks, Gaussian).
The code consistently convolves the extended source shape with the point
spread function (as a function of energy) and integrates the spectrum over energy
to computed model predicted counts.

Finally, \pointlike allows the spatial parameters of an extended
source to be simulataneoulsy fit. \pointlike uses the Minuit fitting
library to perform the maximum likelihood maximization of the position
and extension of an extended source\cite{Minuit reference}.  At each
step in the fit of the shape, the spectral parameters of the entire sky
model are reoptimized. To avoid projection effects, what is directly
fit is not the longitude and latitude of the source but instead
the displacement in a rotated reference frame. The covariance matrix is
then computed and inverted to estimate the statistical errors on the
position and extension.

\pointlike optimizes the analysis of extended sources in two ways.
First, for extended sources which are radially symmetric, the convolution
of the source shape with our parameterization of the PSF can be performed
semi-analytically. Generally, at a given energy the observed 
photon distribution is
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') d A' 
\end{equation}
where $I_\text{src}$ is the extended soruce shape and the PSF is parameterized by
\begin{equation}
  \text{PSF}(\vec r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma}
\end{equation}
Here, $\sigma$ and $\gamma$ are parameterize the $PSF$.
If the extended source is radially symmetric, then
$I_\text{src} (\vec r') = I_\text{src} (r')$ and the angular part of the
integral can be done analytically, leaving a simpler one dimensional
integral:
\begin{equation}
  \text{PDF}(u)= \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma 
  \times ~_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right)
\end{equation}
Not only is this convolution formula faster, but it has the advantage
that it more accurately reduces to the actual PSF shape as the source's
size decreases.  Decreasing the numerical difference between a small extended
source and the PSF is important when searching for smaller extended sources
and when determining the statistical significance of a detection.

@@

To calculate the statistical significance of the extension of a source,
one can define a parameter $\tsext$ as twice the increase
in log likelihood between fitting a source with a
spatially extended hypothesis and a point hypothesis. 
So $\tsext=2\times(LL_\text{ext}-LL_\text{ps})$.

In going from fitting a point source to fitting a simple extended source such as
a disk, the source's width is the only degree of freedom added to the model. According to
Wilk's Theorem, the distribution of $\tsext$ for no signal (when there is a true
point source) should follow a $\chi^2_1$ distribution with one degree of freedom\cite{Wilks_Theorem}.
Wilk's Theorem only holds when the likelihood is maximized simultaneoulsy
fitting all of the parameters of model. It is for this reason why 
\pointlike
goes through such great effort to simultaneously fit the position, extension,
and spectral paramters of a source.

On the other hand, since the null hypothesis rests on the edge of
parameter space, when the extension is fixed to 0, Wilk's Theorem does
not formally hold\cite{Warnings about Wilk's Theorem Paper}.  One might
expect the actual distribution to be a half $\chi^2_1$ distribution with
one degree of freedom plus half of a delta function at 0:
\begin{equation}
  P(\ts)=\tfrac{1}{2}(\chi^2_1(\text{TS})+\delta(\text{TS})).
\end{equation}
The false detection probability can be estimated by integrating this function 
from an observed test statistic value to infinity. It is for this
particular distribution that the often quoted result holds that
$\sqrt{\ts}$ is a measure of the number of $\sigma$ of the detection.

The half $\chi^2_1$ distribution was found by a monte carlo study
to hold for the special case of source finding of EGRET sources using the
maximum likelihood technique.  They were comparing the hypothesis of a
source to the hypothesis of just a background\cite{Mattox_et_All_Paper}.
Their argument for why this null hypothesis distribution was reasonable
was that half of the time, statistical underfluctuations would cause
there to be fewer fewer photons then expected where the sources is being
tested. Since the maximim likelihood machinery cannot fit negative
sources (which would lead to negative model predicted for which the
poisson likelihood is undefined), half of the tests would not have the
likelihood improved by adding a source to the model and $\ts=0$.
It is for the case that the null distribution has

@@

One issue that we found when testing this analysis method is that
despite our best efforts, there was always a small discrepancy between
the likelihood value when fitting the source as a point source and when
fitting it as an extended source with a small extension.  This is to be
expected due to numerical innacuracies introduced by the convolution.
For most cases, this error is insignificant. It would not significanlty
bias any of the fit values or errors. But it would bias the distribution
of $\tsext$ in the null hypothesis. To correct for this,
instead of computing the likelihood in the null hypothesis by fitting
a point source, a new source model was introduced which is an extended
source with its extension fixed to a very small value. $LL_\text{ps}$
is then calculated by refitting this source using the same extended
source fitting code. It was only after introducing this change that we
were able to get broad agreement between simulations and Wilk's theorem.

@@

We performed a dedicated monte carlo study to validate \pointlike for
studying spatially extended sources and also to demonstrate that we can
use $\tsext$ as a test of the statistical significance
of a source's extension.  The monte carlo study involved simulating point
sources of varying spectral models on top of a simulated background. The
simulated point source was then fit using \pointlike as both an extended
source and a point source and $\tsext$ was calculated.

Fermi LAT data was simulated using the publically avaliabable Science
Tools package {\em gtobssim}\cite{GTOBSSIM_CITATION}. The sources were
simulated with a powerlaw spectral model with 6 Were given 6 $>100\mev$
fluxes ranging from $3\times 10^{-9} \text{ph}/\text{cm}^2/\text{s}$ to 
$10^{-6} \text{ph}/\text{cm}^2/\text{s}$.
The sources were then simulated with spectral index ($\gamma$) values of
1.5, 2, 2.5, and 3.  These values were picked to probe a representative
range measurable values.  For particular spectral indices, only the
fluxes which would create a statistically significant source detection
($\ts>25$) where selected, since we are only interested in
testing for extension sources with a significant extent. The sources were
simulated on top of an isotropic powerlaw background with a Sreekumar-like
spectrum ($>100\mev$ flux of 1.5e-5 and spectral index 2.1).\cite{Sreekumar
et al. ApJ 494 pag 523 1998}

The monte carlo simulation was perforormed over a one year time interval using
a default rocking profile and an assumed livetime fraction of 0.8.
The reconstruction was perfromed with events from an energy range of
$1\gev$ to $100\gev$ using 4 energy bins per decade and using the Pass 7 Version 6
source IRFs to be consistent with the following analysis.

@@

The results of the monte carlo simulation are shown in
figure~\ref{ts_ext_mc}.  The cumulative density function (CDF) is plotted
against $\tsext$ For reference, the CDF for $\chi^2_1/2$,
suggested by Wilk's Theorem, is also plotted for comparison. Overall,
there is good agreement between Wilk's theorem, although the agreement
is not perfect.  Possible reasons for departure from Wilk's theorem may
include \pointlike ignoring energy dispersion, which would generally change
the shape as a function of energy, numerical innacuracies in evaluating
the convolution algorithm, or possibly problems with the fitting algorithm
getting stuck in false minimum. It is important to emphasize that if
our emperical curve lies to the left of the curve we assume to estimate
significance, we would underestimate the statistical significance of
our result. Since this is the case for most of our simluations, we are
reasonable confident that $\sqrt{\tsext}$ can be safely
used as a measure of the statistical significance of a source's extension.

@@

We performed another monte carlo study to determine the Fermi LAT's
sensitivity to spatially extended sources. The sensitivity to extension
is defined as the flux at which 
the average value of $\tsext$ for a source is 
e equal to 25 (for a given spectral model). This correseoncs
to an average extension significance of $5\sigma$.  This is not the
sensitivity to detect an extended source but instead the sensitivity to
resolve an already detected source as being extended.

One would intuitivly expect that we would be less sensitivie to resolving
small extended sources since they would be lost in the PSF.  Similarly,
we would be less sensitive to very large extended sources which would
get completely lost in the background. Furtheremore, one expects our
sensitivity to be greater for harder sources since our angular resolution
improves significantly with energy.

To calculate the LAT's sensitivity to spatially extended sources,
we simulated sources with a uniform radially symmetric
disk morphology on top of a powerlaw
Sreekumar-like isotropic spectrum.\cite{Sreekumar et al. ApJ 494 pag
523 1998}.  For a given extension and spectral index, we calculated the
sensitivity by varying the flux of a source until we found 
when $\tsext=25$. In particular, we hand picked a flux range
where $\tsext<25$ and $\tsext>25$ and simulated 10 points with
a log uniform distribution in this range and performed an extension
test for each of these fluxes. Since the measured signifciance is
a statistical which will vary from realization to realization,
we then repeated this procedure >15 times for each realization.
By fitting a line to the flux vs $\tsext$ data, we were able to
estimage the flux at which $\tsext=25$.

Our first result is the extension sensitivity for sources of varying
spectral indices. The result is shown in figure~\ref{index_sensitivity}.
We computed the sensitivty for sources with a spectral index of 1.5,
2, 2.5, and 3 and for sources of 20 extensions varying from $0.1^\circ$
to $2.0^\circ$. The $y$ axis is the $100\mev$ to $100\gev$ flux of the
source and the $x$ axis is the monte carlo extension. The curves then
represent the division in this extension and flux space where one gets
on average $\tsext=25$.
The solid lines of the plot are the sensitivity computed when using
with all events from $100\mev$ to $100\gev$. The dashed lines is the
same sensitivity but only using events from from $1\gev$ to $100\gev$.

This results shows that our extension sensitivity is
significantly better for harder sources where there are more high
energy events. Furthermore, our sensitivity is best for sources with an
extension around $0.5^\circ$.  our sensitivity gets significantly worse
for very small sources and gets slightly worse for very large sources.

Futheremore, we note that except for the hardest or largest sources,
our sensitivity is almost entirly whether or not we use events from
$100\mev$ to $1\gev$. On the other hand, other systematic errors become
increasingly important at low energy so for our extension search we will
not use the events in this energy range.

We performed a second computation to see how the LAT's extension
sensitivity varied with increasing background, as is at lower latitude
in the galactic plane. To do so, we computed the sensitivity using the
extragalactic barground as well as 10 and 100 times the extragalactic
background. The latter result approximatly represents the intensity of
the galactic diffuse emission near the galactic center. The results are
shown in figure~\ref{diff_factor_sensitivity}. Two plots are presented,
one fitting only in the energy range between $1\gev$ and $100\gev$ and
one fitting only in the energy range between $10\gev$ and $100\gev$. For
each plot, the $y$ axis quotes the flux in the fit energy range and the
simulation was done only for a representivite spectral index 2 source.

These plots show that we are much less sensitivie to extended sources in
a higher background region. By comparing the two plots, we also see that
we would be more sensitive to an extended source if it had the same
integral flux, but only over a higher energy range. This is 
consistent with the improved angular resolution at higher energy.

@@

The LAT two year catalog presented a search for all
statistically significant sources that are detected in the $\gev$
sky\cite{2FGL_catalog}.  The catalog group's source finding method assumes
that the emission originated as point sources and what was presented was
a list of all sources fit assuming a point source spatial hyopthesis. All
published extended sources were modeled by the catalog group as extended
sources, but no attempt was made to test if other sources in the catalog
could be spatially resolved.

We performed a search of all Fermi sources found in the second Fermi LAT
catalog for extension.  We performed the extension fit using \pointlike.
To perform this search, we built a model of the sky consistent with
the two year catalog's model.   We used the same two year data set and
the same pass 7 instrument response function We used the same galactic
diffuse, isotropic diffuse, and earth limb emission model. The spectrum
of the galactic and isotropic diffuse models were refit during the
analysis. The earth limb emission was fixed to predicted value.

For each source, we used a circular $10^\circ$ region of interest centered
on our source. We performed a fully binned analysis.  The spatial bin
size was scaled with energy as was described above and we uniformly
binned in log energy, using 4 binned per logarithmic decade of of energy.
We included all background catalog sources within $15^\circ$ of the source
of interest and refit the spectral parameters of sources within $2^\circ$
of our source.

As was shown, we gain very little in sensitivity using events with energy
below $1\gev$. On the other hand, the large PSF at low energy makes us
more succeptable to systmetic errors arising from source confusion and
from out modeling of the galactic diffuse emission.  For that reason,
we performed our search using events between $1\gev$ and $100\gev$.

We also performed a search for extended sources using events only between
$10\gev$ and $100\gev$. Even though we are testing the same sources,
this approach is complimentary because there are regions in the galactic
plane where at lower energies the diffuse emission dominates and is
likely not perfectly modeled. On the other hand, at the highest energies
accessible by the LAT, the galactic diffuse emission will be smaller and
we might expect to detect significantly harder soruces. This approach
of using only events above $10\gev$ was succesfully used to detecting
HESS J1825-137 with the Fermi LAT~\cite{HESS J1825 paper}.

For each source we were testing, we modified the source to have a
uniform radially symmetric disk spatial shape and then used \pointlike to
simultenously fit the position and extension of the source. As described
above, we additionally modified the source to be a extended source with
it's extension fixed to $10^{-10\circ}$ and fit this position of this
pseudo source. For each source, we then computed the increase in log
likelihood fitting with the extended hypothesis and looked at all of
the sources with $\tsext>16$ corresponding to a $4\sigma$ statistical
significance of extension.

We expect most extended sources to be galactic sources since
extragalacitc sources are typically too far away to be spatially
resolved. Unfortuantly, the galactic plane is the most difficult place
to find sources since they must be found for on top of the anisotropic
galactic diffuse emission.  Modeling the galactic diffuse emision is
very difficult~\ref{first_diffuse_paper}.  And finding sources in the
galactic plane is notoriusly difficult, as was discussed in the Fermi
catalog~\ref{1FGL}. One would expect extended sources to be even more
difficult to detect since diffuse emission residuals are expected
to be large compared to the PSF. Furthermore, the galacitc plane
is very crowded and there is a degenecy between an extended source
and multiple point sources. So finding a source with
$\tsext>16$ is not sufficint for detecting an extended.

For each of the extended source candidates for which either the analysis
using events above $1\gev$ or $10\gev$ found $\tsext>16$
and performed several followup analyses to substantiate our analysis.

First, we took the spatial and spectral model found using \pointlike
and performed a crosscheck by refitting the spectral parameters using 
\gtlike.
We used a square ROI $10^\circ$ in diamater, the same 4 degrees
per decade, and $0.125^\circ$ energy binning.
The \gtlike analysis provided a second estimate of the spectral
paramters of the extended source as well as a second calculation of
$\text{TS}_\text{disk}$, $\text{TS}_\text{point}$, and 
$\tsext$. We only looked for extended sources which 
both \pointlike and \gtlike considered significantly extended.

Additionaly, we generated for each 
extended source candidate residual test statistic maps. 
The residual TS Map

Large
TS values in test

We generated a residual test statistic maps to look for
, which is a map of the test
statistic adding new points
Figure~\ref{res_tsmaps} shows an example residual TS map.

\section{Validation of Known Extended Sources}
\label{validate_known}

We validated our method by testing for extension all of the sources
which the Fermi Collaboration had previously published on.  We applied
no special method to test these sources but used the same pipeline which
was used to test all other catalog sources.  The results are published
in table~\ref{known_extended_sources}.  Using events with energies
greater than $1\gev$, our analysis succesfully fit the extension of the
the two galaxies the Large Magalengic Cloud and the Small Magalenic
Cloud.~\cite{SMC & LMC Paper}.  Our tool also successfully detected the
extension of hte supernova remenants IC443, W28, W30, W44, and W51C.
